---
title: "Function names conflict"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, error = TRUE)
```

In RStudio when we load packages into our session we may encounter same function names in different packages. Take e.g. the raster and the dplyr-package (latter part of the tidyverse):

```{r, results = 'as.is', keep  = TRUE}
library(tidyverse)
library(raster)
```

Take careful note of the message you get when loading libraries. It may help when resolving quesitons downstream in your code.

Before we do some raster work we  may first want to do some data manipulations work. E.g. select some variables in the data:

```{r, results = 'as.is'}
minke <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv")
minke %>% 
  select(id, lon, lat)
```

The problem here is that the select function from the dplyr-package is now masked with the select function in the raster-package. And because we loaded the raster-package after the tidyverse-packages (load among other things the dplyr-package) the call is to the function the former. To see the order of your packages you can type:

```{r, results = 'as.is'}
search()
```

Since the raster-package is in enviroment 2 and the dplyr-package in postion 6 any conflicting function name that reside in these two packages will be default be taken from the raster package.

The solution is to add the package-name in front of the function:

```{r}
minke %>% 
  dplyr::select(id, lon, lat)
```

There is a package in development (as of this writing not in R) that may also be helpful, check out [conflicted](https://conflicted.r-lib.org).


