---
title: "Geoinside: Points in polygons via sf"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Lets say we have some spatial areas (polygons) and then some spatial data points. And we are interested to find the area that each point falls into. An example could be finding the ICES statistical rectangles that a set of fishing operations fall under. This problem has been solved many times, this post is only a memo to oneself how this is done using functions in the sf-package using the tidyverse approach.

### Needed libraries

```{r}
library(sf)
library(tidyverse)
```

### A generic case from scratch

Lets start with from scatch creating some 4 rectangle polygons and then a bunch of coordinates representing e.g. fishing haul location which we want to “assign” to each of the four rectangles. More details are provided below, but the final code-flow would be something like:

```{r}
# Generate the 4 rectangles:
icesr <- 
  tibble(Rectangle = c(rep("58C2", 5), rep("58C1", 5), rep("57C2", 5), rep("57C1", 5)),
         lon = c( -28,  -28,  -27,  -27, -28,
                  -29,  -29,  -28,  -28, -27,
                  -28,  -28,  -27,  -27, -28,
                  -29,  -29,  -28,  -28, -27),
         lat = c(64.5, 65.0, 65.0, 64.5, 64.5,
                 64.5, 65.0, 65.0, 64.5, 64.5,
                 64.0, 64.5, 64.5, 64.0, 64.0,
                 64.0, 64.5, 64.5, 64.0, 64.0)) %>% 
  # Generate simple feature POINTS
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326) %>% 
  # Convert to sf MULTIPOINTS, "conditional" on variable Rectangle 
  group_by(Rectangle) %>% 
  summarise(do_union = FALSE) %>% 
  # Convert MULTIPOINTS to POLYGON
  st_cast("POLYGON")

# Generate some random (fishing haul) location
n <- 100
set.seed(314)
haul_location <- 
  tibble(tow = 1:n,
         lon = runif(n, -29.1, -26.9),
         lat = runif(n,  63.9,  65.1)) %>% 
  # here want to keep the lon and the lat as attributes
  mutate(x = lon,
         y = lat) %>% 
  st_as_sf(coords = c("x", "y"),
           crs = 4326)

# Spatial joining
haul_location <- 
  haul_location %>% 
  st_join(icesr["Rectangle"])

# Visualize
ggplot() +
  geom_point(data = haul_location, aes(lon, lat, colour = Rectangle)) +
  geom_sf(data = icesr, alpha = 0.2, aes(fill = Rectangle)) +
  scale_fill_discrete(guide = FALSE) +
  labs(x = NULL, y = NULL)
```

### The final haul data:

```{r}
haul_location
```


I.e. within the haul dataframe the added variable “Rectangle” that each haul falls under. Hauls that are not within any of the four rectangles will have a missing (NA) rectangle variable name.


