---
title: "Polygons and area size"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = TRUE)
```

```{r}
library(tidyverse)
library(units)
library(sf)
```

## Preamble

We have a set of Natura 2000 sites in the North Sea as polygon that overlap. The area of each polygon is known however we are interested in the total area.

## Importing and some tidying

```{r}
d <- 
  read_sf("ftp/shapes/Natura2000_Noordzee1.geojson") 
glimpse(d)
```

**Make variable names a little more "codable" downstream**:
```{r}
d <-  d %>% janitor::clean_names()
glimpse(d)
```

**More pickyness**:

```{r}
d <- 
  d %>% 
  select(id = sitecode,
         name = sitename,
         area_km2 = opp_km)
```

Sanity checks:
```{r}
st_is_valid(d)
```

**Check if area is correct**:
```{r, fig.height = 7}
d <- 
  d %>% 
  mutate(area = st_area(.),
         area = units::set_units(area, km2))
d %>% 
  select(name, area_km2, area) %>% 
  st_drop_geometry()
```

**hmmm, ...**:
```{r, error = TRUE}
ggplot() +
  geom_sf(data = d,
          aes(fill = area), alpha = 0.5)
```

The above could be a bug in the packages, or a bug in ones head. Quick fix:
```{r, error = TRUE}
ggplot() +
  geom_sf(data = d,
          aes(fill = as.numeric(area)), alpha = 0.5)
```

Total sum of the different polygons:

```{r}
sum(d$area)
```

## Solution

But above include the overlaps. To solve that we need to:

* "merge" the polgyons
* then calculate the total area

```{r}
d.union <- 
  d %>% 
  st_union() 
ggplot() +
  geom_sf(data = d.union, alpha = 0.25, fill = "red")

d.union %>% 
  st_area() %>% 
  set_units(km2)
```

