---
title: "Introduction to mapping and spatial analysis with R"
author: ""
output:
  html_document: default
bibliography: mybib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You are here: [https://heima.hafro.is/~einarhj/older/spatialr](https://heima.hafro.is/~einarhj/older/spatialr)

The course announcement is [here](http://www.ices.dk/news-and-events/Training/Pages/R.aspx)

### Time and location
___

__Time__: 2019-11-25 10:00 - 2019-11-29 14:00


__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE, warning=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Preamble
___

During the course we will allocate some time for each participant to work on a personal data, doing some manipulation and producing maps.  Please bring any spatial data that you are interested working with. 

**Prior to the course** please install:

* Latest version of __R__: see [The Comprehensive R Archive Network](https://cran.r-project.org/)
    * If your platform is Windows, also install [Rtools](https://cran.r-project.org/bin/windows/Rtools)

* Latest version of __RStudio__: see [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/)
    * If you are adventurous install the latest [developmental version](https://www.rstudio.com/products/rstudio/download/preview)

For those that want to get a head we recommend the following read:

  * [https://ggplot2-book.org](ggplot2: Elegant Graphics for Data Analysis)
  * [R for Data Science](http://r4ds.had.co.nz) [@wickham2016r]
  * [Robin Lovelace, Jakub Nowosad, Jannes Muenchow: Geocomputation with R](https://geocompr.robinlovelace.net)
  * [Edzer Pebesma, Roger Bivand: Spatial Data Science](https://keen-swartz-3146c4.netlify.com)
  * [Spatial Data Science](https://rspatial.org)
  * [r-spatial](https://www.r-spatial.org)

Additional sites:

  * [stats545](https://stat545.com)
  * [TCRENV2017](https://heima.hafro.is/~einarhj/education/tcrenv2017)
  * [The Visual Raster Cheat Sheet](http://rpubs.com/etiennebr/visualraster)

All the latest things in R:

  * [Rweekly](https://rweekly.org)
  * [Twitter #rstats](https://twitter.com/search?q=%23rstats&src=typed_query)
  * [Twitter #rspatial](https://twitter.com/hashtag/rspatial?src=hashtag_click)

### The workplan
___

(Definitively a) Preliminary schedule:

* Lectures (within the time constraint above): 09:00 - 17:00
* Own work, we will be around: 17:00 - 18:00
* Lunch (definite): 13:00 - 14:00
* Coffee (plan): 11:00 - 11:20, 15:30 - 15:45

#### Day 1 - Monday (10:00 - 17:00)

* [R & RStudio](pre_intro.html) (EH)
* [Spatial data in R](pre_spatial_data.html) (JB)
* [Introduction to ggplot](pre_ggplot1.html) (EH)

#### Day 2 - Tuesday (09:00 - 17:00)

* [Data transformation](pre_transformation.html) (EH)
* [The sf class](pre_sf.html) (JB)
* [Reading/writing spatial data](pre_input_output.html)(JB)

#### Day 3 - Wednesday (09:00 - 17:00)

* [Coordinate Reference Systems](pre_crs.html) (JB)
* [Working with geometries](pre_geometric.html) (JB)
* [Reproducible document writing in R?](pre_markdown.html) (EH) 
* [Spatial operations](pre_spatial_ops.html) (JB)


#### Day 4 - Thursday (09:00 - 17:00)

* [Rasters](pre_rasters.html) (JB)
* [Interactive maps - leaflet](pre_interactive.html) (EH)

#### Day 5 - Friday (09:00 - 13:00)

* [Interpolation](pre_interpolation.html) (JB)
* [Little more on ggplot2](pre_ggplot2.html) (EH)
* Recap
* Course evaluation and certificates


## Bits and pieces
___

* [Categorising continuos variable](bit_category.html)
* [Downloading data](bit_download.html)
* [Some details on raster merging](bit_raster.html)
* [Function name conflicts](bit_conflicts.html)
* [Points in polygons](bit_points_in_polygon.html)
* [Time series graphs - visualizing vms data](bit_time_series_graphs.html)
* [Merging VMS and logbooks](bit_logs_and_vms.html) - solving using time
* [Popups in leaflet](bit_leaflet_popups.html)

### Datasets used

* [Information of datasets used - clean up after the course](datasets_overview.html)

### References
___

Web page source is found at [https://github.com/fishvice/spatialr](https://github.com/fishvice/spatialr).
