---
title: "Introduction to mapping and spatial analysis with R"
author: ""
output:
  html_document: default
bibliography: mybib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

You are here: [https://heima.hafro.is/~einarhj/spatialr](https://heima.hafro.is/~einarhj/spatialr)

### Time and location
___

__Time__: 2019-11-25 10:00 - 2019-11-29 15:00

__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE, warning=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Preamble
___

**Prior to the course** please install:

* Latest version of __R__: see [The Comprehensive R Archive Network](https://cran.r-project.org/)
    * If your platform is Windows, also install [Rtools](https://cran.r-project.org/bin/windows/Rtools)

* Latest version of __RStudio__: see [RStudio Desktop](https://www.rstudio.com/products/rstudio/download/)
    * If you are adventurous install the latest [developmental version](https://www.rstudio.com/products/rstudio/download/preview)

For those that want to get a head we recommend the following read:

  * [R for Data Science](http://r4ds.had.co.nz) [@wickham2016r]
  * [Robin Lovelace, Jakub Nowosad, Jannes Muenchow: Geocomputation with R](https://geocompr.robinlovelace.net)
  * [Edzer Pebesma, Roger Bivand: Spatial Data Science](https://keen-swartz-3146c4.netlify.com)

### The workplan
___

(Definitively a) Preliminary schedule:

#### Day 1 - Monday

* [Introduction](pre_intro.html) - eh
* [Introduction: Plotting spatial data](pre_ggplot2.html) - eh

#### Day 2 - Tuesday

* 
* 


#### Day 3 - Wednesday

* 
* 
* [Exercises using DATRAS data](pre_case.html) - eh

#### Day 4 - Thursday

* 
* 


#### Day 5 - Friday

* 
* 

### Bits and pieces
___


### Potential projects
___


### References
___

Web page source is found at [https://github.com/fishvice/spatialr](https://github.com/fishvice/spatialr).
