---
title: "Introduction to ggplot2"
subtitle: "Part 1"
output:
  html_document:
    fig_height: 4
    fig_width: 9
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
    toc_depth: 2
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Getting started
___

**Needed libraries**:
```{r}
library(tidyverse)   # installs sweeps of packages
library(patchwork)   # used for multiplotting
library(maps)        # some basic country maps
library(mapdata)     # higher resolution maps
library(marmap)      # access global topgraphy data
```

```{r, echo = FALSE}
theme_set(theme_grey(base_size = 16))
cl = function(x, color = "blue"){
  outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{", x ,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>", x ,"</font>",sep="")
  else
    x
}
```

**Importing external data**:

In this tutorial we are going to use the minke data and the Icelandic spring groundfish survey data (see full explonation of variables in [Datasets - overview](datasets_overview.html)).

```{r}
minke <- 
  read.csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble()
glimpse(minke)
```

```{r}
smb <- 
  read.csv("ftp://ftp.hafro.is/pub/data/csv/is_smb.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble() %>% 
  filter(year == 2019)
smb %>% select(id:cod_n) %>% glimpse()
```

## Key components of ggplot 

ggplot has __three__ key components: 

1.  argument __data__, which must be a `data.frame` or some derivative there of (`tbl`, `data.table`, ...)

2. A set of aesthetic mappings (called with the `r cl("aes")`~function) between variables in the data and visual properties, and 

3. At least one call to a `r cl("layer")` which describes how to render each observation.


```{r}
ggplot(data = minke) +
  aes(x = lon, y = lat) + 
  layer(geom = "point", stat = "identity", position = "identity")
```

Generally we do not call `r cl("layer")` directly but use functions starting with `r cl("geom_")` that are a shortcut calls to the `r cl("layer") `-function. Hence the above call is normally written as:
```{r, eval = FALSE}
ggplot(data = minke, aes(x = lon, y = lat)) + geom_point()
```

Different syntax, equivalent outcome:

```{r, eval = FALSE}
ggplot()                        + geom_point(data = minke, aes(lon, lat))
ggplot(data = minke)            + geom_point(aes(x = lon, y = lat))
ggplot(minke)                   + geom_point(aes(lon, lat))
ggplot(minke, aes(lon, lat))    + geom_point()
```

A ggplot can be stored as an object for later use:

```{r}
p <- ggplot(minke) + geom_point(aes(lon, lat))
```

In the above plot, we basically have mapped the longitude position on the x-axis and the latitude position on the y-axis. There are two things missing:

* A background or a reference, giving the reader a better indication of the geographical region of the sample location.
* The projection looks odd.

## Reference map

* Maps as background for r-plot can come from myriad of sources. Here we take an example of shoreline that reside in the map-packages.
* To get the data into required ggplot2 form (a data.frame) we use the `r cl("map_data")`-function.

```{r}
iceland <- map_data("worldHires", region = "Iceland")
glimpse(iceland)
```

Here we have just a simple dataframe with `r nrow(iceland)` coordinates (long and lat) and some other variables. We can try map these coordinates to different layers (more on `r cl("labs")` later):

```{r}
m <- ggplot(iceland, aes(long, lat, group = group)) + labs(x = NULL, y = NULL)
m1 <- m + geom_point()
m2 <- m + geom_line()
m3 <- m + geom_path()
m4 <- m + geom_polygon()
m1 + m2 + m3 + m4
```

The above sweep of code and plots demonstrate that background maps are just a set of longitudinal and latitudinal data that are **ordered** in a specific way.

## Projections

A proper coverage of projections is done in a [xxx](xxx.hmtl), here we just cover some basics with respect to ggplot2.

As noted above a map is just a xy-plot but with a certain projections. We could try to guess the projections (or rather the aspect ration of the plot) as done on the left or better use the specific functions available in ggplot:

```{r}
m1 <- m3 + coord_fixed(ratio = 2.4) 
m2 <- m3 + coord_quickmap()
m3 <- m3 + coord_map()
m4 <- m3 + coord_map(projection = "stereographic")
```

```{r, echo = FALSE}
m1 <- m1 + annotate("text", x = -18, y = 64.8, label = "ration: 2.4")
m2 <- m2 + annotate("text", x = -18, y = 64.8, label = "coord_quickmap")
m3 <- m3 + annotate("text", x = -18, y = 64.8, label = "coord_map")
m4 <- m4 + annotate("text", x = -18, y = 64.8, label = "stereographic")
```

```{r, fig.height = 5}
m1 + m2 + m3 + m4
```

Note that the `r cl("geom_quickmap")` is an approximation (albeit good one for most purposes), if one is operating on a fine scale `r cl("geom_map")` may be more accurate (actually all maps are wrong when put on a two dimensional pane).

## Adding layers

The power of ggplot comes into place when one adds layers on top of other layers. With the following code one first generates the background-map and then adds the minke datapoints on top of that background map:

```{r}
p <-
  ggplot() +
  geom_polygon(data = iceland, 
               aes(long, lat, group = group),
               fill = "grey") +
  geom_point(data = minke, 
             aes(lon, lat, colour = sex)) +
  labs(x = NULL, y = NULL)
p + coord_map()
```

Besides adding a `r cl("point")` layer on **top** of the `r cl("polygon")` layer we have added two aesthetics:

* `fill`: This argument overrides the default fill argument in ggplot ("black"). Take note that it is called outside the `r cl("aes")`-function (so it acts like a constant).
* `colour`: Here we are letting the value of the variable sex control the colour displayed. Hence we call that argument within the `r cl("aes")`-function.

We will be introduced to more aesthetics as we move along.

## Zooming in on a plot

We can zoom into a plot by specifying boundaries of the horizontal and vertical axis by calling the arguments xlim and ylim within the `r cl("coord_map")`-function.

```{r}
p + coord_map(xlim = c(-25, -20), ylim = c(65, 66.6))
```

## Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* Emulate the following map of the Faeroe Islands (or use whatever country you like) using the data in the map-package (hint: Faeroe Island is a colony of a country).
* What happens if you do not use the group-argument?

```{r, echo = FALSE}
map_data("worldHires", region = "Denmark") %>% 
  ggplot() +
  geom_polygon(aes(long, lat, group = group),
               fill = "grey", colour = "darkgreen") +
  labs(x = NULL, y = NULL) +
  coord_map(xlim = c(-9, -5), ylim = c(61.25, 62.5))
```

</div>
</div>

# Layers
___

Here we will only focus on ggplot `r cl("layers")` that are useful when plotting spatial data.

Since we are going to use some Icelandic survey data in the following example (and for the sake of code clarity) we first generate a reference background layer:
```{r}
m <- 
  ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL)
```

The `r cl("scale_*_continuous")`-function is used here to suppress the default name and breaks that ggplot assigns (this would not need to be suppressed in R base-plot).

## geom_segment

The `r cl("segment")`-layer is useful when wanting to display some very simple vector data that only have a start and an end point. Lets take the tow location from the Icelandic spring survey as an example:

```{r}
m +
  geom_segment(data = smb,
               aes(x    = lon1, y    = lat1,
                   xend = lon2, yend = lat2))
```

## geom_point

We can assign a variable to the size argument in the `r cl("geom_point")`-layer, the size of the point then being some function of value of a variable. Lets visualize the abundance of cod:

```{r}
p <-
  m +
  geom_point(data = smb,
             aes(lon1, lat1, size = cod_n),
             alpha = 0.2,
             colour = "red")
p
```

Here we have introduced a new aesthetic named alpha. This argument controls the transparency of the bubble, the value going from 0 (invisible) to 1 (no transparency, the ggplot default).

Besides the spatial patterns observed, there is one problem with this plot: The radius of the bubble size is not proportional to abundance. ggplot provides a very useful function to remedy this (now zero abundance gets a zero radius):

```{r}
p +
  scale_size_area(max_size = 20)
```

The value assigned to max_size is a bit of a trial and error as well as personal preference.

### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* Create a code that emulates the following plot for the abundance of cod (or for any other variable of interest).

```{r, echo = FALSE}
ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  geom_segment(data = smb,
               aes(x    = lon1, y    = lat1,
                   xend = lon2, yend = lat2)) +
  geom_point(data = smb,
             aes(lon1, lat1, size = cod_n),
             alpha = 0.2,
             colour = "red") +
  scale_size_area(max_size = 20)
```

</div>
</div>

## geom_path

We have already shown a plot using the `r cl("path")`-layer, but just to create a background map. For a more appropriate demonstration lets use the track data from the 2019 spring survey. Besides the coordinate data the key additional element of this dataset is time, which it has already been ordered by.

```{r}
# move this to the header (or not)
track <- 
  read.csv("ftp://ftp.hafro.is/pub/data/csv/is_smb_vms2019.csv")
```

```{r}
m +
  geom_path(data = track,
            aes(x = lon, y = lat, colour = vessel))
```

## geom_tile

The survey data contains the ICES rectangles that a tow belongs (ir) to as well as the center coordinate position (ir_lon and ir_lat) we can use the `r cl("tile")`-layer to display some characteristics of the data (we will later deal with how this was achieved).

However, since there is more than one tow in each tile we need first to generate some summary of the data by each rectangle (actually here by the center-points):

```{r}
s <-
  smb %>% 
  group_by(ir_lon, ir_lat) %>% 
  summarise(Cod = mean(cod_n))
p <-
  m +
  geom_tile(data = s,
            aes(x = ir_lon, y = ir_lat, fill = Cod))
p
```

Do not like the default colour?. Try:
```{r, eval = FALSE}
p +
  scale_fill_viridis_c(option = "B", direction = -1)
```

This display has still at least two problems:

* The background reference (Iceland) is now hidden behind some tiles (see exercise below).
* The mapping of the value on the colour scale used may not be the most reviling (much more details on that later).


#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Ameliorate the plot above such that is looks like this (you have to start from scratch):

```{r, echo = FALSE}
s <-
  smb %>% 
  group_by(ir_lon, ir_lat) %>% 
  summarise(Cod = mean(cod_n))
ggplot() +
  geom_tile(data = s,
            aes(x = ir_lon, y = ir_lat, fill = Cod)) +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_viridis_c(option = "B", direction = -1)
```

</div>
</div>

## geom_raster

The above plot using the `r cl("tile")`-layer is actually a kind of a raster plot, albeit the resolution being not very high.

To demonstrate the use of the `r cl("raster")`-layer lets use the global relief models from the ETOPO1 dataset hosted on a NOAA server can be accessed using the `getNOAA.bathy`-function in the marmap-package (you need to install that package). To access them one specifies the boundary of the data of interest and then, since we are using ggplot for mapping are turned into a data frame using the `r cl("fortify")`-function:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)

depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1) %>% 
  fortify() %>%  # turn the object into a data.frame
  filter(z <= 0)
glimpse(depth)
```

So this data is just a set of x (longitude), y (latitudes) and z (depth). The dataset is a raster-grid which we can visualize by using the `geom_raster`-function:

```{r}
m +
  geom_raster(data = depth,
              aes(x = x, y = y, fill = z)) +
  coord_quickmap(expand = FALSE) +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  labs(fill = "Depth [m]")
```

## geom_contour

We will later deal with the syntax for generating contours from rasters. However, as is often the case, ggplot provides a shortcut for generating contours based on raster data, the `r cl("conour")`-layer.

```{r}
m2 <- 
  m +
  geom_contour(data = depth, aes(x, y, z = z),
               breaks = c(-25, -50, -100, -200, -400),
               lwd = 0.1) 
m2
```

Here we have specified to display the depth contours of 25, 50, 100, 200 and 400 meters.

Now we are ready to add the minke data or any other data of interest:

```{r, eval = FALSE}
m2 + geom_point(data = minke, aes(lon, lat), colour = "red")
```


#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Create a depth raster map of the region of your interest.
2. Create a contour map of the region of your interest, specifying your own preference for the depth values to display.

and / or

3. Create a code that emulates the following plot

```{r}
ggplot() +
  geom_raster(data = depth,
              aes(x = x, y = y, fill = z)) +
  geom_contour(data = depth,
               aes(x = x, y = y, z = z),
               breaks = -400,
               colour = "white") +
  geom_segment(data = smb,
               aes(x = lon1, y = lat1,
                   xend = lon2, yend = lat1),
               colour = "cyan") +
  geom_point(data = smb,
             aes(x = lon1, y = lat1, size = cod_age1),
             colour = "red",
             alpha = 0.5) +
  scale_size_area(max_size = 10) +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  coord_quickmap()
```

</div>
</div>
