---
title: "Introduction to ggplot2"
output:
  html_document:
    fig_height: 4
    fig_width: 9
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
    toc_depth: 2
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Getting started
___

## Libraries

**Needed libraries**:
```{r}
library(tidyverse)   # installs sweeps of packages
library(maps)        # some basic country maps
library(mapdata)     # higher resolution maps
library(marmap)      # access global topgraphy data
```

```{r, echo = FALSE}
library(patchwork)   # used for multiplotting
```

```{r, echo = FALSE}
theme_set(theme_grey(base_size = 16))
cl = function(x, color = "blue"){
  outputFormat = knitr::opts_knit$get("rmarkdown.pandoc.to")
  if(outputFormat == 'latex')
    paste("\\textcolor{",color,"}{", x ,"}",sep="")
  else if(outputFormat == 'html')
    paste("<font color='",color,"'>", x ,"</font>",sep="")
  else
    x
}
```

## Key components of ggplot 

**Data used**:
```{r}
minke <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/minke.csv")
glimpse(minke)
```

See full explanation of variables in [Datasets - overview](datasets_overview.html)).

ggplot has __three__ key components: 

1.  argument __data__, which must be a `data.frame` or some derivative there of (`tbl`, `data.table`, ...)

2. A set of aesthetic mappings (called with the `r cl("aes")`~function) between variables in the data and visual properties, and 

3. At least one call to a `r cl("layer")` which describes how to render each observation.


```{r}
ggplot(data = minke) +
  aes(x = lon, y = lat, colour = sex) + 
  layer(geom = "point", stat = "identity", position = "identity")
```

Generally we do not call `r cl("layer")` directly but use functions starting with `r cl("geom_")` that are a shortcut calls to the `r cl("layer") `-function. Hence the above call is normally written as:
```{r, eval = FALSE}
ggplot(data = minke, aes(x = lon, y = lat, colour = sex)) + geom_point()
```

Different syntax, equivalent outcome:

```{r, eval = FALSE}
ggplot()                        + geom_point(data = minke, aes(lon, lat, colour = sex))
ggplot(data = minke)            + geom_point(aes(x = lon, y = lat, colour = sex))
ggplot(minke)                   + geom_point(aes(lon, lat, colour = sex))
ggplot(minke, aes(lon, lat, colour = sex))    + geom_point()
```

A ggplot can be stored as an object for later use:

```{r}
p <- ggplot(minke) + geom_point(aes(lon, lat, colour = sex))
```

In the above plot, we basically have mapped the longitude position on the x-axis and the latitude position on the y-axis. There are two things missing:

* A background or a reference, giving the reader a better indication of the geographical region of the sample location.
* The projection looks odd.

## Reference map

* Maps as background for r-plot can come from myriad of sources. Here we take an example of shoreline that reside in the map-packages.
* To get the data into required ggplot2 form (a data.frame) we use the `r cl("map_data")`-function.

```{r}
iceland <- map_data("worldHires", region = "Iceland")
glimpse(iceland)
```

Here we have just a simple dataframe with `r nrow(iceland)` coordinates (long and lat) and some other variables. We can try map these coordinates to different layers (more on `r cl("labs")` later):

```{r, fig.height = 5}
m <- 
  ggplot(iceland, aes(long, lat, group = group))
m1 <- m + geom_point()
m2 <- m + geom_line()
m3 <- m + geom_path()
m4 <- m + geom_polygon()
```

```{r, echo = FALSE}
m1 <- m1 + annotate("text", x = -18, y = 64.8, label = "point", colour = "red")
m2 <- m2 + annotate("text", x = -18, y = 64.8, label = "line", colour = "red")
m3 <- m3 + annotate("text", x = -18, y = 64.8, label = "path", colour = "red")
m4 <- m4 + annotate("text", x = -18, y = 64.8, label = "polygon", colour = "red")
```

```{r, echo = FALSE, fig.height = 6}
m1 + m2 + m3 + m4
```

The above sweep of code and plots demonstrate that background maps are just a set of longitudinal and latitudinal data that are **ordered** in a specific way.

## Projections

As noted above a map is just a xy-plot but with the **addition** of having a projections. A proper coverage of projections is done in a [xxx](xxx.hmtl), here we just cover some basics with respect to ggplot2. We could try to guess the projections (or rather the aspect ration of the plot) as done on the left or better use the specific functions available in ggplot:

```{r}
m <- 
  ggplot(iceland, aes(long, lat, group = group)) +
  geom_path()
m1 <- m + coord_fixed(ratio = 2.4) 
m2 <- m + coord_quickmap()
m3 <- m + coord_map()
m4 <- m + coord_map(projection = "stereographic")
```

```{r, echo = FALSE}
m1 <- m1 + annotate("text", x = -18, y = 64.8, label = "ration: 2.4")
m2 <- m2 + annotate("text", x = -18, y = 64.8, label = "coord_quickmap")
m3 <- m3 + annotate("text", x = -18, y = 64.8, label = "coord_map")
m4 <- m4 + annotate("text", x = -18, y = 64.8, label = "stereographic")
```

```{r, echo = FALSE, fig.height = 6}
m1 + m2 + m3 + m4
```

Note that the `r cl("geom_quickmap")` is an approximation (albeit good one for most purposes), if one is operating on a larger scale `r cl("geom_map")` may be more accurate (actually all maps are wrong when put on a two dimensional pane).

## Adding layers

The power of ggplot comes into place when one adds layers on top of other layers. With the following code one first generates the background-map and then adds the minke datapoints on **top** of that background map:

```{r}
m <-
  ggplot() +
  geom_polygon(data = iceland, 
               aes(long, lat, group = group)) +
  geom_point(data = minke, 
             aes(lon, lat, colour = sex))
m + coord_quickmap()
```

Besides adding a `r cl("point")` layer on **top** of the `r cl("polygon")` layer we have added one aesthetics:

* `colour`: Here we are letting the value of the variable sex control the colour displayed. Hence we call that argument within the `r cl("aes")`-function.

We will be introduced to more aesthetics as we move along.

## Customization

ggplot2 provides a lot of sensible defaults that it derives from the data used to generate a plot. Here we will take a peek on those of that may be of most interest for now, further customization being introduced as we move along.

When one writes:

```{r}
p <- 
  ggplot(minke) + 
  geom_point(aes(lon, lat, colour = sex))
```

what ggplot does is actually:

```{r, eval = FALSE}
ggplot(minke) + 
  geom_point(aes(lon, lat, colour = sex)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

Within the `r cl("scale_")`-functions all the settings take the default argument, check e.g.:
```{r, eval = FALSE}
?scale_x_continuous
```

Here we observe among other things: name, breaks, minor_breaks, limits, position, ... Here is an example where among other the default name and breaks are overwritten:

```{r}
p + 
  scale_x_continuous(name = "Longitude",
                     breaks = seq(-25, -10, by = 1),
                     minor_breaks = NULL) +
  scale_y_continuous(name = "Latitude",
                     breaks = seq(63, 67, by = 0.5),
                     minor_breaks = NULL) +
  scale_colour_discrete(name = "Sex") +
  coord_map()
```

If one was interested in:

* Overwrite the default black colour in the polygon with grey (fill = "grey")
* Suppressing the axis labels (name = NULL)
* Suppressing the axis values (breaks = NULL)
* Change the default colour palette to brewer scale (palette = "Set1")
* Place the legend in the center of the figure (theme(legend.position = c(0.5, 0.5)))

we would do:
```{r}
m <- 
  ggplot() +
  geom_polygon(data = iceland, 
               aes(long, lat, group = group),
               fill = "grey") +
  geom_point(data = minke,
             aes(lon, lat, colour = sex)) +
  scale_x_continuous(name = NULL,
                     breaks = NULL) +
  scale_y_continuous(name = NULL,
                     breaks = NULL) +
  scale_colour_brewer(name = "Sex",
                      palette = "Set1") +
  theme(legend.position = c(0.5, 0.5)) +
  coord_map()
m
```

If one only interested in overwriting the default axis-name one could use the `r cl("labs")`-shortcut function.

One of ggplot defaults is that the range of the x and the y-axis is dictated by the range in the data. If we want to "zoom into" a plot we can overwrite the default boundaries of the horizontal and vertical axis by calling the arguments xlim and ylim within the `r cl("coord_quickmap")`-function.

```{r}
m + coord_quickmap(xlim = c(-25, -20), ylim = c(65, 66.6))
```

## Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* Emulate the following map of the Faeroe Islands (or use whatever country you like) using the data in the map-package (hint: Faeroe Island is a colony of a country).
* What happens if you do not use the group-argument?

The minke data used is:

```{r}
minke.fo <- 
  tibble(lon = c(-8.5, -5.5),
         lat = c(61.5, 62.2),
         sex = c("Male", "Female"))
```

```{r, echo = FALSE}
fo <- map_data("worldHires", region = "Denmark")

ggplot() +
  geom_polygon(data = fo,
               aes(long, lat, group = group),
               fill = "grey", colour = "darkgreen") +
  geom_point(data = minke.fo,
             aes(lon, lat, colour = sex)) +
  scale_x_continuous(breaks = seq(-9, -5, by = 0.5)) +
  scale_color_brewer(palette = "Set1") +
  labs(x = NULL, y = NULL) +
  coord_map(xlim = c(-9, -5), ylim = c(61.25, 62.5))
```

</div>
</div>

# More layers
___

Here we will only focus on ggplot `r cl("layers")` that are useful when plotting spatial data.

**Data used**:
```{r}
smb <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb.csv") 
smb2019 <- filter(smb, year == 2019)
glimpse(smb2019)
```

Since we are going to use some Icelandic survey data in the following example we first generate an appropriate reference background layer:
```{r}
m <- 
  ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL)
```

## geom_segment

The `r cl("segment")`-layer is useful when wanting to display some very simple vector data that only have a start and an end point. Lets take the tow location from the Icelandic spring survey as an example:

```{r}
m +
  geom_segment(data = smb2019,
               aes(x    = lon1, y    = lat1,
                   xend = lon2, yend = lat2))
```

## geom_point

We can assign a variable to the size argument in the `r cl("geom_point")`-layer, the size of the point then being some function of value of a variable. Lets visualize the abundance of cod:

```{r}
p <-
  m +
  geom_point(data = smb2019,
             aes(lon1, lat1, size = cod_n),
             alpha = 0.2,
             colour = "red")
p
```

Here we have introduced a new aesthetic named alpha. This argument controls the transparency of the bubble, the value going from 0 (invisible) to 1 (no transparency, the ggplot default).

Besides the spatial patterns observed, there is one problem with this plot: The radius of the bubble size is not proportional to abundance. ggplot provides a very useful function to remedy this (now zero abundance gets a zero radius):

```{r}
p +
  scale_size_area(max_size = 20)
```

The value assigned to max_size is a bit of a trial and error as well as personal preference.

### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* Create a code that emulates the following plot for the abundance of cod (or for any other variable of interest).

```{r, echo = FALSE}
ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  geom_segment(data = smb2019,
               aes(x    = lon1, y    = lat1,
                   xend = lon2, yend = lat2)) +
  geom_point(data = smb2019,
             aes(lon1, lat1, size = cod_n),
             alpha = 0.2,
             colour = "red") +
  scale_size_area(max_size = 20)
```

</div>
</div>

## geom_path

We have already shown a plot using the `r cl("path")`-layer, but just to create a background map. For a more appropriate demonstration lets use the vessel track data from the 2019 spring survey. Besides the coordinate data the key additional element of this dataset is time, which it has already been ordered by.

```{r}
track <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb_vms2019.csv")
```

```{r}
m +
  geom_path(data = track,
            aes(x = lon, y = lat, colour = vessel))
```

## geom_tile

The survey data contains the ICES rectangles that a tow belongs (ir) to as well as the center coordinate position (ir_lon and ir_lat) we can use the `r cl("tile")`-layer to display some characteristics of the data (we will later deal with how this was achieved).

However, since there is more than one tow in each tile we need first to generate some summary of the data by each rectangle (actually here by the center-points):

```{r}
s <-
  smb2019 %>% 
  group_by(ir_lon, ir_lat) %>% 
  summarise(Cod = mean(cod_n))
p <-
  m +
  geom_tile(data = s,
            aes(x = ir_lon, y = ir_lat, fill = Cod))
p
```

Do not like the default colour?. Try:
```{r, eval = FALSE}
p +
  scale_fill_viridis_c(option = "B", direction = -1)
```

This display has still at least two problems:

* The background reference (Iceland) is now hidden behind some tiles (see exercise below).
* The mapping of the value on the colour scale used may not be the most reviling (much more details on that later).


#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Ameliorate the code for the plot above such that is looks like this (you have to start from scratch):

```{r, echo = FALSE}
s <-
  smb2019 %>% 
  group_by(ir_lon, ir_lat) %>% 
  summarise(Cod = mean(cod_n))
ggplot() +
  geom_tile(data = s,
            aes(x = ir_lon, y = ir_lat, fill = Cod)) +
  geom_polygon(data = iceland, aes(long, lat, group = group),
               fill = "grey") +
  coord_map() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_viridis_c(option = "B", direction = -1)
```

</div>
</div>

## geom_raster

The above plot using the `r cl("tile")`-layer is actually a kind of a raster plot, albeit the resolution being not very high.

To demonstrate the use of the `r cl("raster")`-layer lets use the global relief models from the ETOPO1 dataset hosted on a NOAA server can be accessed using the `getNOAA.bathy`-function in the marmap-package (you need to install that package). To access them one specifies the boundary of the data of interest and then, since we are using ggplot for mapping are turned into a data frame using the `r cl("fortify")`-function:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)

depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1) %>% 
  fortify() %>%  # turn the object into a data.frame
  filter(z <= 0)
glimpse(depth)
```

So this data is just a set of x (longitude), y (latitudes) and z (depth). The dataset is a raster-grid which we can visualize by using the `geom_raster`-function:

```{r}
m +
  geom_raster(data = depth,
              aes(x = x, y = y, fill = z)) +
  coord_quickmap() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  labs(fill = "Depth [m]")
```

## geom_contour

We will later deal with the syntax for generating contours from rasters. However, as is often the case, ggplot provides a shortcut for generating contours based on raster data, the `r cl("contour")`-layer.

```{r}
m2 <- 
  m +
  geom_contour(data = depth, aes(x, y, z = z),
               breaks = c(-25, -50, -100, -200, -400),
               lwd = 0.1) 
m2
```

Here we have specified to display the depth contours of 25, 50, 100, 200 and 400 meters.

Now we are ready to add the minke data or any other data of interest:

```{r, eval = FALSE}
m2 + geom_point(data = minke, aes(lon, lat), colour = "red")
```


#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Create a depth raster map of the region of your interest.
2. Create a contour map of the region of your interest, specifying your own preference for the depth values to display.

and / or

3. Create a code that emulates the following plot

```{r, echo = FALSE}
ggplot() +
  geom_raster(data = depth,
              aes(x = x, y = y, fill = z)) +
  geom_contour(data = depth,
               aes(x = x, y = y, z = z),
               breaks = -400,
               colour = "white") +
  geom_segment(data = smb2019,
               aes(x = lon1, y = lat1,
                   xend = lon2, yend = lat1),
               colour = "cyan") +
  geom_point(data = smb2019,
             aes(x = lon1, y = lat1, size = wolffish_kg),
             colour = "red",
             alpha = 0.5) +
  scale_size_area(max_size = 10) +
  # do not show legend for the depth raster
  scale_fill_continuous(guide=FALSE) +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  coord_quickmap()
```

</div>
</div>

# Facetting

Facetting allows one to split up the data by a variable and display the same graph for each subset. One can split the data by one variable (wrap) or by two variables (grid).

## facet_wrap

Lets say we want to visualize the survey biomass of a species (here haddock) over time. For simplicity here we will only use selected 6 years:

```{r}
smb2 <- 
  # more on filter later
  filter(smb, year %in% c(1995, 2000, 2005, 2010, 2015, 2019))
p <-
  m + 
  geom_point(data = smb2,
             aes(lon1, lat1, size = haddock_kg),
             alpha = 0.15,
             colour = "red") +
  scale_size_area(max_size = 15)
p +
  facet_wrap(~ year)
```

## facet_grid

**data used**: Survey abundance data of cod by year, age and yearclass at each station.

```{r}
rbya <- 
  read_csv("ftp://ftp.hafro.is/pub/data/csv/is_smb_cod_rbya.csv")
rbya <- filter(rbya, yc %in% c(1984, 1985, 1996, 1997))
glimpse(rbya)
```

Note: Here we have filtered out four year classes.

```{r, fig.height = 7}
m + 
  geom_point(data = rbya,
             aes(lon1, lat1, size = n),
             alpha = 0.2,
             colour = "red") +
  scale_size_area(max_size = 10) +
  facet_grid(age ~ yc) +
  labs(title = "Abundance of cod", 
       subtitle = "By yearclass and age",
       size = "per 4 miles")
```

Note: The in the `r cl("facet_grid")`-function the first variable (here age) is plotted as rows, the second variable (here yearclass) as columns.

