---
title: "Working with geometries in sf objects"
output:
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: cerulean
    toc: yes
    toc_float: yes
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### These are the packages needed for this tutorial:
```{r setup}
library(tidyverse)
library(sf)
```

## Operating with single geometries in sf object
The package **sf** provides a large set of functions to query properties of sf objects and to combine sf objects to form new geometries.

Geometric operations (and the **sf** functions that implement them) can be classified as:

- predicates: they return a logical value (TRUE or FALSE)
- measures: they return a value
- geometry generator operations: they return new geometries.

Operations can also be classified as unary, binary, or n-ary, depending if they need one, two or more *sf* objects.

#### Remember to use projected data
- We want to get areas in m2...not square degrees!
- For geometric operations:
  - Make sure that they are in a *projected* crs.
  - Make sure that all layers have the *same* crs.

## Extracting basic information
**sf** has several functions to extract basic information from objects of class *sf*:

```{r }
nfu <- read_sf("ftp://ftp.hafro.is/pub/data/shapes/nephrops_fu.gpkg") %>%
 st_transform(3395) # Transform to World Mercator

st_is_simple(nfu)

st_is_valid(nfu)

st_is_empty(nfu)

st_is_longlat(nfu)

st_dimension(nfu) # 0=points, 1=linear, 2=polygon

st_geometry_type(nfu)

st_crs(nfu) # Query (or set) the coordinate reference system

st_bbox(nfu) # Bounding  box

```

## Generating new geometries

#### Sampling
The function **st_sample()** can be used to sample locations from polygons or linestrings.  The result is an *sfc* object with the sampled POINT geometries.

```{r }

nfu <- read_sf("ftp://ftp.hafro.is/pub/data/shapes/nephrops_fu.gpkg") %>%
 st_transform(3395) # Transform to World Mercator

set.seed(100)
pts1 <- st_sample(nfu, size = 100, type = "random")

nfu_lines <- st_cast(nfu, "LINESTRING")
pts2 <- st_sample(nfu_lines, size = 100, type = "random")

ggplot() +
  geom_sf(data = nfu) +
  geom_sf(data = pts1, color = "red") +
  geom_sf(data = pts2, color = "blue")

```

#### Centroids
The mean position of geometries (LINESTRING OR POLYGONS) can be obtained by the *st_centroid()* function.

```{r }
centr <- st_centroid(nfu)
centr2 <- st_point_on_surface(nfu) # Not exactly the centroid, but always in the polygon

ggplot() +
  geom_sf(data = nfu) +
  geom_sf(data = centr, color = "red") +
  geom_sf(data = centr2, color = "blue")

```

#### Buffers
The function *st_buffer()* can be used for points, linestrings or polygons.

```{r }
buf <- st_buffer(pts1, dist = 100000) # 100 km buffer
ggplot() +
  geom_sf(data = buf, linetype = "dashed") +
  geom_sf(data = pts1)

buf <- st_buffer(nfu_lines, dist = 50000)
ggplot() +
  geom_sf(data = buf, aes(fill = name))


buf <- st_buffer(nfu, dist = 20000)
ggplot() +
  geom_sf(data = buf, aes(fill = name))

# Negative buffers are allowed for polygons!
buf <- st_buffer(nfu, dist = -20000)
ggplot() +
  geom_sf(data = buf, aes(fill = name))

```
#### Convex hull
A convex hull is the smallest convex polygon that includes a group of points.  We can use the *st_convex_hull()* function to get it, but first we need to join the POINT geometries into a single MULTIPOINT geometry.  Otherwise we get a "convex hull" for each individual point.

```{r }

pts1_u <- st_union(pts1)

chull <- st_convex_hull(pts1_u)

ggplot() +
  geom_sf(data = chull) +
  geom_sf(data = pts1)

```

*st_union()*  can be used to simplify features.  For polygons, it will combine multiple polygons into a single polygon.  For points, it clusters individual points into a MULTIPOINT geometry.

#### Concave hulls
Concave hulls are similar to convex hull, but they are concave (duh!).  To compute a concave hull we need the package **concaveman**  Note that the *concaveman* takes an *sf* object, but it does not accept an *sfc* object. (i.e. a geometry)

```{r }
library(concaveman)

conc1 <- concaveman(st_as_sf(pts1), concavity = 1)
conc2 <- concaveman(st_as_sf(pts1), concavity = 2.5)

ggplot() +
  geom_sf(data = conc1, color = "red", fill= NA) +
  geom_sf(data = conc2, color = "blue", fill = NA) +
  geom_sf(data = pts1)

```
With high values for the *concavity* parameter we get a convex hull.

#### Grids
The function *st_grid* provides rectangular or hexagonal grids.  You can get polygons, nodes or centroids.

```{r }

sq_grid <- st_make_grid(nfu, n = c(10, 15))

hex_grid <- st_make_grid(nfu, n = c(10, 15), square = FALSE)

ggplot() +
  geom_sf(data = sq_grid, fill = NA, color = "gray") +
  geom_sf(data = hex_grid, fill = NA, color = "red") +
  theme_bw()

```
Sometimes we want to make a polygon of the bounding box of a geometry.  For this, we can make a grid with a single cell.

```{r }
box <- st_make_grid(nfu, n = 1)

```
The *st_graticule()* can be used to create a set of lines with constant latitude or longitude, which can be added to maps as reference.

#### Simplification

Sometimes it is necessary to simplify vector data, reducing the memory, disk space and bandwidth they require, and speeding up geometrical operations.  For this we use the *st_simplify()* function.

```{r }

helcom <- read_sf("ftp://ftp.hafro.is/pub/data/shapes/helcom.gpkg") %>%
  st_transform(3395)

helcom_simple <- st_simplify(helcom, dTolerance = 5000)

object.size(helcom)
object.size(helcom_simple)

ggplot() + geom_sf(data = helcom)
ggplot() + geom_sf(data = helcom_simple)

```


## Measurements
#### Areas and lengths
The **sf** package provides a series of unary measures that return a single values describing some properties of geometries.

```{r }

st_area(nfu)

st_length(nfu) # Polygons have no length!

st_length(nfu_lines)


```
#### Distances between objects
The function *st_distance()* returns a dense numeric matrix with distances between geometries.

Let's get a few points, lines and polygons:

```{r }
set.seed(100)
poly <- nfu[1:3, ]

pts_a <- st_sample(poly, 5) %>%
  st_as_sf()%>%
  mutate(lab = 1:5)

pts_b <- st_sample(poly, 2)

st_distance(pts_a, pts_b)

st_distance(pts_a, poly)

ggplot() +
  geom_sf(data = poly, aes(fill = name)) +
  geom_sf(data = pts_a) +
  geom_sf_text(data = pts_a,
               nudge_x = 20000,
               aes(label = lab))


```
The function returns a distance matrix with all pairwise distances (but see the *by_element* argument).  The matrix is of class *units*, for numerical data with their associated measurement unit.

Note that we get a distance of 0 when the point is inside the polygon.



## Aggregation
Aggregation involves summarising a dataset by a grouping variables, usually one of the attribute columns.  The *sf* package provides methods for stats::aggregate and dplyr::summarise. When used, they:

- take a groping predicate (e.g. from *group_by()*)
- take one or more aggregation functions
- aggregate the attributes per group
- aggregate the geometries
- if *do_union==TRUE*, union the aggregated geometries.

Let's look again to our small vms dataset:

```{r }
vms <- read_csv("ftp://ftp.hafro.is/pub/data/csv/small_vms.csv") %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

vms # It is geometry type POINT.

vms_gr <- vms %>%
  group_by(id) %>%
  summarise(n = n())

vms_gr # It is geometry MULTIPOINT!

vms_tr <- vms %>%
  group_by(id) %>%
  summarise(n = n()) %>%
  st_cast("LINESTRING")

vms_tr # It is now a LINESTRING.
```

Note that the order of the points in the LINESTRING changed.  This is because by default the geometries are combined using *st_union()*  In this case though it is better to use *st_combine(), by adding * do_union = FALSE*.

```{r }
plot(vms_tr %>% st_geometry())

vms_tr <- vms %>%
  group_by(id) %>%
  summarise(n = n(),
            do_union = FALSE) %>%
  st_cast("LINESTRING")

plot(vms_tr %>% st_geometry())

```

Let's aggregate the ICES ecoregions into our three groups:


```{r }
ices_er <- read_sf("ftp://ftp.hafro.is/pub/data/shapes/ices_ecoregions.gpkg") %>%
   mutate(zone = c(1, 2, 2, 3, 3, 3, 3, 3, 2, 2, 2, 1, 1, 1, 1, 1, 2)) # Add a grouping variable: 1=Arctic & subarctic, 2=North Atlantic, 3=Mediterranean & Black Sea

ices_er_zones <- ices_er %>%
  group_by(zone) %>%
  summarize(area = sum(area_km2))

ggplot() +
  geom_sf(data = ices_er_zones,
          aes(fill = area))

```

When aggregating polygons, it is better to allow the default *do_union=TRUE* to dissolve internal boundaries.  Otherwise we get invalid polygons (and internal divisions in our plots).

We can use rbind to join spatial objects, but the column names and coordinate reference system need to be the same.

```{r }
sm_ices_er <- ices_er %>%
  filter(area_km2 < 100)

lg_ices_er <- ices_er %>%
  filter(area_km2 >= 100)

all_ices_er <- rbind(sm_ices_er, lg_ices_er)

```
## Relationship between attributes and geometry
Most **sf** objects contain attributes and geometry.  Because geometric operations do not change existing attributes, we need to be careful about if the attributes *make sense* in terms of the new geometries.

Example:

- We take some trawls within a polgyon representing our study area.
- We calculate the total abundance of some species (say cod) in that polygon.  This is an attribute of the polygon (POP_COD)
- We split the polygon into three areas based on bottom depth.
- The attribute POP_COD is assigned to the three depthbased polygons... but it is not "right" anymore.

We need to distinguish between three attributegeometry relationships (AGR)

1. **Constant attributes**.  They are valid everywhere within a geometry.  Examples: bottom type, depth class.
2. **Aggregate attributes**. They are a summary value over the geometry.  Examples: total abundance, mean abundance.
3. **Identity attributes**. They identify uniquely the *entire* geometry.  Example: EEZs, ICES ecoregions.  A sample of this geometry has not the identity property anymore.  It becomes a constant attribute.

When we operate with geometries, the **sf** package gives a warning that attributes assumed to be constant over a geometry (which may be true or not).

```{r }
st_centroid(ices_er) %>%
  head(1)

```
We can specify the AGR like this:

```{r }
st_agr(ices_er) # None of the variables have a set AGR.

st_agr(ices_er) <- c(ecoregion = "identity",
                     area_km2 = "aggregate",
                     zone = "constant")

st_agr(ices_er) # Now they do.

ices_ct <- st_centroid(ices_er)

st_agr(ices_er)
st_agr(ices_ct)

```

