---
title: ""
output:
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
```

# Data
___

#### Needed libraries for this tutorial:
```{r, message = FALSE, warning = FALSE, results='hide'}
library(sf)
library(leaflet)
library(tidyverse)
theme_set(theme_grey(base_size = 16))
```

## Minke

The minke whale dataset contains biological measurements from 192 scientific catches of minke whales between the year 2003 and 2007. The data set contains the following variables:

* **whale.id**: Unique identifier for the whale
* **date.caught**: the date when the whales was caught
* **lat**: latitude
* **lon**: longitude
* **area**: Derived from location (North/South)
* **length**: length of the whale
* **weight**: weight of the whale
* **age**: age of the whale
* **sex**: Male or Female
* **maturity**: maturity status of the whale
* **stomach.volume**: volume (in liters) of the stomach content
* **stomach.weight**: weight of the stomach content
* **year**: the year when the whale was caught

```{r}
minke <- 
  read.csv("http://www.hafro.is/~einarhj/data/minke.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble()
```


## IESSNS 2019 survey

### trail

Dataset contains VMS tracks of 5 vessels that participated in IESSNS 2019 survey (often named the Pelagic Mackerel Survey). The variables are:

* **vessel**: The vessel name
* **time**: Time in UTC
* **lon**: Degrees longitude
* **lat**: Degrees latitute
* **speed**: Vessel speed in knots
* **heading**: Vessel heading in dregrees

```{r}
trail <- 
  read.csv("ftp://ftp.hafro.is/pub/reiknid/einar/data/iessns2019/trail.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble()
trail %>% glimpse()
```

### tows

Dataset contains pre-determined pelagic trawl stations. The variables are:

* **activity**: Some information on the type of tow
* **lon**: Degrees longitude
* **lat**: Degrees latitute

```{r}
tows <- 
  read.csv("ftp://ftp.hafro.is/pub/reiknid/einar/data/iessns2019/tows.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble()
tows %>% glimpse()
```

### strata

Dataset contains the survey stratas (used when calculating the stratified abundance/biomass indices). The variables are:

* **lon**: Degrees longitude
* **lat**: Degrees latitute
* **strata**: Numerical indicating strata number

```{r}
strata <- 
  read.csv("ftp://ftp.hafro.is/pub/reiknid/einar/data/iessns2019/strata.csv",
           stringsAsFactors = FALSE) %>% 
  as_tibble()
strata %>% glimpse()
```